---
title: "SD Tutorial 6: S-Shaped Growth — SIS Epidemic Model"
subtitle: "Two interacting stocks, opposing flows, and the dynamics of infection"
author: "Kenneth Moore"
format:
  html:
    number-sections: false
    toc: true
    theme: united
    highlight-style: tango
---

```{r}
#| label: setup
#| include: false

library(deSolve)
library(tidyverse)

source("../R/sd-diagram-functions.R")
```

**Required R packages**
```{r}
#| label: packages
#| eval: false

library(deSolve)
library(tidyverse)
```

# Acknowledgement

The core content behind this tutorial derives from the [Road Maps](https://ocw.mit.edu/courses/sloan-school-of-management/15-988-system-dynamics-self-study-fall-1998-spring-1999/readings/) curriculum, developed by the System Dynamics in Education Project (SDEP) at MIT under the direction of Professor Jay Forrester.

Source material for this tutorial comes from [Generic Structures: S-Shaped Growth I](https://ocw.mit.edu/courses/sloan-school-of-management/15-988-system-dynamics-self-study-fall-1998-spring-1999/readings/) (D-4432-2), Section: S-Shaped Growth Structure 2.

# S-Shaped Growth Is a Behaviour, Not a Structure

In [Tutorial 5](tut-05-s-shaped-growth.qmd) we saw S-shaped growth emerge from a **single stock** with a density-dependent deaths multiplier. This tutorial shows that the same characteristic behaviour can arise from a completely different mechanism — **two interacting stocks** with no lookup tables at all.

The key insight: S-shaped growth is a *behaviour pattern*, not a *structure*. Multiple mechanistically distinct systems can produce it.

# The SIS Epidemic Structure

This model describes the spread of an illness through a fixed population of 100 people. There is no permanent immunity: once a sick person recovers, they return to the healthy pool and can be re-infected. This is known as an **SIS** (Susceptible → Infected → Susceptible) structure.

The two stocks are:

- **Healthy People** — those susceptible to infection
- **Sick People** — those currently infected

Two flows connect them in opposing directions:

- **Catching Illness** — moves people from Healthy to Sick; driven by contact between healthy and sick people
- **Recovering** — moves people from Sick back to Healthy; driven by the duration of illness

```{r}
#| label: diagram
#| echo: false
#| fig-align: center
#| fig-width: 8
#| fig-height: 4.5

flow_arrow <- arrow(length = unit(0.44, "cm"), type = "closed")
info_arrow <- arrow(length = unit(0.32, "cm"), type = "open")

ggplot() +
  # Healthy People stock
  annotate("rect", xmin = 0.3, xmax = 4.3, ymin = 2.5, ymax = 3.5,
           fill = "white", colour = "black", linewidth = 1.6) +
  annotate("rect", xmin = 0.42, xmax = 4.18, ymin = 2.62, ymax = 3.38,
           fill = "#fafaf0", colour = "black", linewidth = 0.5) +
  annotate("text", x = 2.3, y = 3.0, label = "Healthy\nPeople",
           size = 3.5, fontface = "bold") +
  # Sick People stock
  annotate("rect", xmin = 6.7, xmax = 10.7, ymin = 2.5, ymax = 3.5,
           fill = "white", colour = "black", linewidth = 1.6) +
  annotate("rect", xmin = 6.82, xmax = 10.58, ymin = 2.62, ymax = 3.38,
           fill = "#fafaf0", colour = "black", linewidth = 0.5) +
  annotate("text", x = 8.7, y = 3.0, label = "Sick\nPeople",
           size = 3.5, fontface = "bold") +
  # Catching Illness flow (rightward, blue)
  annotate("segment", x = 4.3, xend = 6.7, y = 3.28, yend = 3.28,
           linewidth = 2, colour = "#4e8cd4", arrow = flow_arrow) +
  annotate("text", x = 5.5, y = 3.68, label = "Catching\nIllness",
           size = 3.2, colour = "#4e8cd4") +
  # Recovering flow (leftward, red)
  annotate("segment", x = 6.7, xend = 4.3, y = 2.72, yend = 2.72,
           linewidth = 2, colour = "#d45b4e", arrow = flow_arrow) +
  annotate("text", x = 5.5, y = 2.42, label = "Recovering",
           size = 3.2, colour = "#d45b4e") +
  # Auxiliary: prob. contact with sick (top centre, above Catching Illness)
  geom_polygon(data = make_ellipse(5.5, 4.65, rx = 1.6, ry = 0.42), aes(x, y),
               fill = "white", colour = "grey50", linewidth = 0.6) +
  annotate("text", x = 5.5, y = 4.65, label = "prob. contact\nwith sick",
           size = 3.0) +
  # Info: Healthy People → prob. contact with sick (curved up and right)
  geom_curve(aes(x = 2.3, y = 3.5, xend = 3.9, yend = 4.65),
             curvature = -0.3, linetype = "solid", colour = "grey40",
             arrow = info_arrow) +
  # Info: Sick People → prob. contact with sick (curved up and left)
  geom_curve(aes(x = 8.7, y = 3.5, xend = 7.1, yend = 4.65),
             curvature = 0.3, linetype = "solid", colour = "grey40",
             arrow = info_arrow) +
  # Info: prob. contact with sick → Catching Illness (straight down)
  annotate("segment", x = 5.5, xend = 5.5, y = 4.23, yend = 3.9,
           linetype = "solid", colour = "grey40", arrow = info_arrow) +
  # Loop labels
  annotate("text", x = 2.5, y = 4.65, label = "B1  (-)", size = 3.8,
           colour = "#c0392b", fontface = "bold") +
  annotate("text", x = 8.7, y = 4.65, label = "R1  (+)", size = 3.8,
           colour = "#27ae60", fontface = "bold") +
  theme_void() +
  coord_cartesian(xlim = c(0, 11), ylim = c(0.3, 5.2)) +
  labs(title = "SIS Epidemic — stock-and-flow diagram")
```

Two feedback loops drive this system:

- **R1 (+)**: More sick people → more probability of contact with sick people → more probability of catching illness → more sick people. This reinforcing loop accelerates the epidemic.
- **B1 (-)**: More healthy people → lower probability of contact with sick people → less catching illness → fewer sick people → fewer people recovering → fewer healthy people. This balancing loop controls disease spread over time.


# Model Equations

The catching illness rate depends on three factors: the number of healthy people available to be infected, the probability of contact with a sick person (sick/total population), and the rate and likelihood of transmission on contact.

$$\text{Catching Illness} = \text{Healthy} \times \frac{\text{Sick}}{\text{Total Population}} \times \text{Population Interactions} \times P(\text{catching})$$

$$\text{Recovering} = \frac{\text{Sick}}{\text{Duration of Illness}}$$

The differential equations governing the two stocks are:

$$\frac{d(\text{Healthy})}{dt} = \text{Recovering} - \text{Catching Illness}$$

$$\frac{d(\text{Sick})}{dt} = \text{Catching Illness} - \text{Recovering}$$

Note that `total population = Healthy + Sick` is conserved — the model has no births or deaths, only movement between compartments.

## Parameters

| Parameter | Symbol | Value | Units |
|-----------|--------|-------|-------|
| Duration of illness | $D$ | 0.5 | months |
| Population interactions | $PI$ | 10 | contacts / person / month |
| Probability of catching | $PC$ | 0.5 | dimensionless |

# R Modelling Tutorial

## Model function

```{r}
#| label: model

model <- function(time, stocks, params) {
  with(as.list(c(stocks, params)), {

    a_total_population  <- s_healthy + s_sick
    a_prob_contact_sick <- s_sick / a_total_population

    f_catching_illness <- s_healthy * a_prob_contact_sick *
                          p_population_interactions * p_prob_catching   # people/month
    f_recovering    <- s_sick / p_duration_of_illness                # people/month

    ds_healthy_dt <- f_recovering - f_catching_illness
    ds_sick_dt    <- f_catching_illness - f_recovering

    return(list(c(ds_healthy_dt, ds_sick_dt),
                catching_illness = f_catching_illness,
                recovery_rate    = f_recovering))
  })
}
```

## Parameters and simulation time

```{r}
#| label: params

params <- c(p_duration_of_illness     = 0.5,   # months
            p_population_interactions = 10,    # contacts/person/month
            p_prob_catching           = 0.5)   # dimensionless

simtime <- seq(0, 5, by = 0.125)               # months
```

## Single test run

Before running multiple scenarios, let us verify the model produces sensible output for one set of initial conditions: 10 sick people and 90 healthy people.

```{r}
#| label: test-run

stocks_a <- c(s_healthy = 90, s_sick = 10)

test_run <- ode(y = stocks_a, times = simtime, func = model,
                parms = params, method = "rk4") %>%
  as_tibble() %>%
  pivot_longer(-time) %>%
  mutate(value = as.numeric(value),
         time  = as.numeric(time))

test_run %>%
  filter(name %in% c("s_healthy", "s_sick")) %>%
  ggplot(aes(x = time, y = value, colour = name)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c(s_healthy = "#4e8cd4", s_sick = "#d45b4e"),
                      labels = c(s_healthy = "Healthy People",
                                 s_sick    = "Sick People")) +
  labs(title  = "SIS epidemic — single run (10 sick, 90 healthy)",
       x      = "Months",
       y      = "People",
       colour = NULL)
```

The system reaches a stable equilibrium rather than infecting the entire population. The epidemic overshoots slightly then settles — a classic S-shaped trajectory for sick people.

## Exercise 2: four initial conditions

To understand how initial conditions affect behaviour, we run four scenarios matching Exercise 2 from the Road Maps source material. We wrap the simulation in a function and use `purrr::pmap_dfr()` to iterate across all four scenarios at once.

```{r}
#| label: wrapper-function

run_sis <- function(sim_id, init_healthy, init_sick,
                    start = 0, finish = 5, step = 0.125) {
  stocks <- c(s_healthy = init_healthy, s_sick = init_sick)
  sim_time <- seq(start, finish, by = step)

  ode(y = stocks, times = sim_time, func = model,
      parms = params, method = "rk4") %>%
    as_tibble() %>%
    pivot_longer(-time) %>%
    mutate(value = as.numeric(value),
           time  = as.numeric(time),
           run   = sim_id)
}
```

```{r}
#| label: four-runs

sim_results <- pmap_dfr(
  list(
    sim_id       = c("a: 5 sick, 95 healthy",
                     "b: 0 sick — no epidemic",
                     "c: 55 sick, 45 healthy",
                     "d: 50 sick, 50 healthy"),
    init_healthy = c(95, 100, 45, 50),
    init_sick    = c(5,   0, 55, 50)
  ),
  .f = run_sis
)

glimpse(sim_results)
```

```{r}
#| label: plot-four-runs

sim_results %>%
  filter(name %in% c("s_healthy", "s_sick")) %>%
  mutate(name = recode(name,
                       s_healthy = "Healthy People",
                       s_sick    = "Sick People")) %>%
  ggplot(aes(x = time, y = value, colour = run)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~name) +
  labs(title  = "Exercise 2 — four initial conditions",
       x      = "Months",
       y      = "People",
       colour = "Run")
```

The four runs reveal the key behaviours of this system:

- **Run a (5 sick)** and **run d (50 sick)**: both converge to the same equilibrium of 40 healthy / 60 sick, regardless of where they started. The system has a single stable attractor.
- **Run c (55 sick)**: starts above the equilibrium sick count and decays back down to it — the epidemic recedes.
- **Run b (0 sick)**: with no sick people to start, the positive feedback loop cannot ignite. Total population stays healthy — this is an unstable equilibrium at `Sick = 0`.

## Run e: terminal illness

What happens if there is no recovery? A terminal illness removes the balancing loop B1 entirely. Without recovery pulling people back to the healthy pool, the epidemic follows a pure S-curve until the entire population is sick.

```{r}
#| label: model-no-recovery

model_no_recovery <- function(time, stocks, params) {
  with(as.list(c(stocks, params)), {

    a_total_population  <- s_healthy + s_sick
    a_prob_contact_sick <- s_sick / a_total_population

    f_catching_illness <- s_healthy * a_prob_contact_sick *
                          p_population_interactions * p_prob_catching
    f_recovering    <- 0

    ds_healthy_dt <- -f_catching_illness
    ds_sick_dt    <-  f_catching_illness

    return(list(c(ds_healthy_dt, ds_sick_dt),
                catching_illness = f_catching_illness,
                recovery_rate    = f_recovering))
  })
}
```

```{r}
#| label: run-e

simtime_e <- seq(0, 8, by = 0.125)   # extended to show full S-curve

run_e <- ode(y = c(s_healthy = 90, s_sick = 10), times = simtime_e,
             func = model_no_recovery, parms = params, method = "rk4") %>%
  as_tibble() %>%
  pivot_longer(-time) %>%
  mutate(value = as.numeric(value),
         time  = as.numeric(time))

run_e %>%
  filter(name %in% c("s_healthy", "s_sick")) %>%
  mutate(name = recode(name,
                       s_healthy = "Healthy People",
                       s_sick    = "Sick People")) %>%
  ggplot(aes(x = time, y = value, colour = name)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("Healthy People" = "#4e8cd4",
                                 "Sick People"     = "#d45b4e")) +
  labs(title  = "Run e — terminal illness (no recovery)",
       x      = "Months",
       y      = "People",
       colour = NULL)
```

Without recovery, sick people grow in a classic S-shape — slow at first (few sick), then rapid (positive loop dominates), then levelling off as the healthy pool is exhausted (negative loop B2 dominates). The system reaches a new equilibrium of `Sick = 100, Healthy = 0`.

# Equilibrium

At equilibrium, inflow equals outflow for both stocks:

$$\text{Catching Illness} = \text{Recovering}$$

Substituting the model equations:

$$H^* \times \frac{S^*}{N} \times PI \times PC = \frac{S^*}{D}$$

Where $N = H^* + S^* = 100$ is the fixed total population. Dividing both sides by $S^*$ (assuming $S^* \neq 0$) and rearranging:

$$H^* = \frac{N}{PI \times PC \times D} = \frac{100}{10 \times 0.5 \times 0.5} = \frac{100}{2.5} = 40$$

Therefore:

$$S^* = N - H^* = 100 - 40 = 60$$

The system always settles at **40 healthy, 60 sick** — regardless of initial conditions — as long as $S_0 > 0$. This equilibrium is confirmed by all the runs above (a, c, and d all converge to the same values).

```{r}
#| label: verify-equilibrium

sim_results %>%
  filter(name %in% c("s_healthy", "s_sick"),
         time == max(time)) %>%
  select(run, name, value) %>%
  mutate(value = round(value, 2)) %>%
  pivot_wider(names_from = name, values_from = value)
```

# Conclusion

This tutorial demonstrated S-shaped growth emerging from a two-stock SIS epidemic structure. Several insights stand out:

1. **S-shaped growth is a behaviour, not a structure.** The same S-curve seen in the single-stock density model (Tutorial 5) arises here from a fundamentally different mechanism — the interaction between two populations.

2. **The product of two stocks creates nonlinearity.** The catching illness rate is proportional to `Healthy × (Sick/Total)`, which is what generates the characteristic S-shape without any lookup table.

3. **Equilibrium can be derived algebraically.** Setting catching illness equal to Recovering and solving gives $H^* = N / (PI \times PC \times D)$ — a clean formula for the stable endemic level.

4. **Initial conditions determine trajectory, not destination.** All positive-$S_0$ runs converge to the same equilibrium; only the path differs.

5. **Removing recovery eliminates the stable interior equilibrium.** Without recovery (run e), the only equilibrium is full infection — confirming that B1 is essential for containing the epidemic.
